\documentclass{article}
\input{preamble.tex}

\begin{document}

\title{Notes on Bi-Quadratic Extensions}

\author{Ken Lee}
\date{Spring 2021}
\maketitle

%\tableofcontents

\begin{prop}[Classification of Bi-Quadratic Extensions]

  Let $a, b \in K$ a field of characteristic not $2$ and 
  $b$ non-square in $K$.
  Let $K \to L$ be a splitting field of $(X^2 - a)^2 - b$.
  The roots are then $\pm \sqrt{a \pm \sqrt{b}}$ and we have the tower,
  \begin{cd}
      &
        & K(\sqrt{a + \sqrt{b}}) \ar[rd]
          & \\
    K \ar[r,"2"]
      & K(\sqrt{b}) \ar[ru] \ar[rd]
        & 
          & L \\
      &
        & K(\sqrt{a - \sqrt{b}}) \ar[ru]
          &
  \end{cd}
  Define the mysterious $c := a^2 - b$.
  Then the following are true : 
  \begin{enumerate}
    \item Let $c$ be a square in $K$. Then we have cases : 
    \begin{enumerate}
      \item Let either $2(a + \sqrt{c}), 2(a - \sqrt{c})$ square in $K$.
      (We cannot have both since $b$ non-square in $K$.)
      Then $L = K(\sqrt{b})$ and hence $\AUT_K L \iso C_2$.
      \item Let both $2(a \pm \sqrt{c})$ non-squares in $K$.
      Then $[L : K] = 4$ and $\AUT_K L \iso C_2 \times C_2$.
    \end{enumerate}
    \item Let $c$ non-square in $K$. Then we have cases : 
    \begin{enumerate}
      \item Let $bc$ square in $K$. Then $[L : K] = 4$ and $\AUT_K L \iso C_4$.
      \item Let $bc$ non-square in $K$.
      Then $[L : K] = 8$ and $\AUT_K L \iso D_{2(4)}$.
    \end{enumerate}
  \end{enumerate}
\end{prop}
\begin{proof}
  Throughout the proof,
  we use $\al, \al'$ to denote $\sqrt{a + \sqrt{b}}, \sqrt{a - \sqrt{b}}$
  respectively. 
  We first give an ansatz of what $\si \in \AUT_K L$ could do 
  by exploiting $\AUT_K L \hookrightarrow \AUT\set{\pm\al,\pm\al'}$.
  Let $\si \in \AUT_K L$ which we view as a subgroup of the permutations of 
  the roots. We have the cases : 
  \begin{itemize}
    \item $\si(\al) = \al$. Then $\si(-\al) = -\al$.
    So we have two cases : 
    \begin{itemize}
      \item $\si = \id{}$
      \item $\si = (\al' -\al')$, a reflection.
    \end{itemize}
    \item $\si(\al) = \al'$. Then $\si(-\al) = -\al'$.
    So we have two cases : 
    \begin{itemize}
      \item $\si = (\al \,\,\al') (-\al \,\, -\al')$, a diagonal reflection.
      \item $\si = (\al \,\, \al' \,\, -\al \,\, -\al')$.
    \end{itemize}
    \item $\si(\al) = -\al$. Then we have either : 
    \begin{itemize}
      \item $\si = (\al \,\, -\al)$, a reflection.
      \item $\si = (\al \,\, -\al) (\al' \,\, -\al')
      = (\al \,\, \al' \,\, -\al \,\, -\al')^2$.
    \end{itemize}
    \item $\si(\al) = -\al'$. Then either : 
    \begin{itemize}
      \item $\si = (\al \,\, -\al') (-\al \,\, \al')$, a diagonal reflection.
      \item $\si = (\al \,\, -\al' \,\, -\al \,\, \al')
      = (\al \,\, \al' \,\, -\al \,\, -\al')\inv$
    \end{itemize}
  \end{itemize}
  So we see $\AUT_K L \iso D_{2(4)}$ when its cardinality is largest,
  otherwise we obtain subgroups of $D_{2(4)}$.

  The Galois group $\AUT_K L$ depends crucially on two things : 
  \begin{enumerate}
    \item [(Q1)] Are $K(\sqrt{b}) \to K(\al)$, 
    $K(\sqrt{b}) \to K(\al')$ trivial?
    \item [(Q2)] Is $K(\al) = K(\al')$?
  \end{enumerate} 
  Considering (Q1), let $a \pm \sqrt{b}$ is a square in $K(\sqrt{b})$.
  Then we have $x, y \in K$ such that 
  \begin{align*}
    a \pm \sqrt{b} 
    &= (x + y \sqrt{b})^2 = x^2 + b y^2 + 2 xy \sqrt{b} 
    \implies a = x^2 + b\brkt{\frac{\pm 1}{2x}}^2 \\
    \implies 0 &= x^4 - a x^2 + b/4 
    \implies x^2 = \frac{a \pm \sqrt{c}}{2}
  \end{align*}
  where $c := a^2 - b$, the discriminant of the quadratic in $x^2$,
  which is not so mysterious afterall.

  \textit{(Assume $c$ square in $K$)}
  \textit{(One of $2(a \pm \sqrt{c})$ square in $K$)}
  Suppose $2(a + \sqrt{c})$ square in $K$.
  Then $K(\al) = K(\sqrt{b})$.
  Since $\al \al' = \sqrt{c} \in K$, 
  we hence have $L = K(\al,\al') = K(\sqrt{b})$.
  As for the Galois group, 
  since $[L : K] = 2$, we cannot have any 4-cycles in $\AUT_K L$.
  This leaves the identity, 
  the reflections and $(\al \,\, -\al) (\al' \,\, -\al')$.
  Any automorphism fixing $\sqrt{b}$ or $\al$ or $\al'$ 
  must be identity since $L = K(\sqrt{b}) = K(\al) = K(\al')$.
  This leaves the identity and 
  two reflections diagonal reflections.
  Finally, which reflection is in $\AUT_K L$ hinges on 
  which of $\al + \al'$ or $\al - \al'$ is in $K$.
  This is the other significance of $2(a + \sqrt{c})$ : 
  $(\al + \al')^2 = 2(a + \sqrt{c})$.
  So by assumption $\al + \al' \in K$.
  Now if $(\al \,\,-\al') (\al' \,\, -\al) \in \AUT_K L$,
  we would have $\al - \al' \in K$ and hence $\al \in K$, a contradiction.
  Thus, $\AUT_K L$ is generated by $(\al \,\,\al') (-\al \,\, -\al')$.

  The above argument can be repeated instead with assuming 
  $2(a - \sqrt{c})$ is a square in $K$.
  Then everything is the same until the last step, 
  where we find $\AUT_K L$ is generated by $(\al \,\, -\al') (-\al \,\, \al')$
  instead. 

  \textit{(Both $2(a \pm \sqrt{c})$ non-square in $K$)}
  Both $a \pm \sqrt{b}$ are not squares in $K(\sqrt{b})$,
  so $K(\sqrt{b}) \to K(\al), K(\sqrt{b}) \to K(\al')$ are both degree $2$.
  But again $\al \al' = \sqrt{c} \in K$, 
  so $L = K(\al,\al') = K(\al) = K(\al')$ and hence $[L : K] = 4$.
  Any automorphism fixing $\al$ or $\al'$ must be identity,
  so this excludes the non-diagonal reflections.
  Since $(\al \pm \al')^2 = 2(a \pm \sqrt{c})$,
  we have at least three sub-$K$-extensions $K(\sqrt{b}),K(\al \pm \al')$.
  By Galois theory, this gives at least three subgroups of $\AUT_K L$
  with index $2$.
  It thus cannot be the case that $\AUT_K L$ is generated by 
  the 4-cycles,
  and hence $\AUT_K L$ is generated by the diagonal reflections,
  isomorphic to $V_4$.

  \textit{($c$ non-square in $K$)}
  The extensions $K(\sqrt{b}) \to K(\al), K(\sqrt{b}) \to K(\al')$ must 
  be order $2$.
  The question is $(Q2)$.
  After thinking for a while, 
  a thing one might try is the following : 
  \begin{lem}
    Let $F$ be a field with non-even characteristic.
    Let $r,s \in F$ with $s$ non-square in $F$.
    Then $\sqrt{r} \in F(\sqrt{s})$ if and only if 
    $rs$ or $r$ is a square in $F$.
    \begin{proof1}
      $(\implies)$
      Let $r = (x + y\sqrt{s})^2$ with $x,y \in F$.
      We get $r = x^2 + sy^2$ and $0 = 2 xy$. 
      So either $r = x^2$ or $rs = (s y)^2$.

      $(\limplies)$
      $r = x^2 / s = (x / \sqrt{s})^2$. 
    \end{proof1}
  \end{lem}
  So we have $\al' \in K(\al) = K(\sqrt{b})(\al)$ if and only if 
  $c = \al \al'$ or $a - \sqrt{b}$ square in $K(\sqrt{b})$.
  The latter implies $(a - \sqrt{c})/2$ is a square in $K$
  and hence $\sqrt{c} \in K$, a contradiction.
  So $K(\al') = K(\al)$ if and only if $c$ is a square in $K(\sqrt{b})$.
  But this is if and only if $\sqrt{c} \in K(\sqrt{b})$,
  which by another application of the lemma,
  is equivalent to $bc$ or $c$ being a square in $K$.
  The latter is again false by assumption,
  so we have $K(\al) = K(\al')$ if and only if $bc$ is a square in $K$.

  \textit{($bc$ is a square in $K$)}
  We have $L = K(\al) = K(\al')$ and $[L : K] = 4$.
  Again, any automorphism fixing $\al$ or $\al'$ must be identity 
  so we are left with the rotations and the diagonal reflections.
  If $\AUT_K L$ is generated by the diagonal reflections,
  then $[K(\al \pm \al') : K]$ would be order two and hence 
  $\min(\al\pm\al',K) = X^2 - 2(a \pm \sqrt{c})$,
  in particular $\sqrt{c} \in K$, a contradiction. 
  So $\AUT_K L$ is generated by a 4-cycle.

  \textit{($bc$ is non-square in $K$)} Covered.
\end{proof}

%\printbibliography

\end{document}